Gravitational Lens Recovery with \Glass: How to measure the mass profile and shape of a lens

### 2013-06-25 limits/ms.tex
More edits.

\begin{abstract}
We use a new non-parametric gravitational -- \Glass -- to determine what quality of data (strong lensing, stellar kinematics, and/or stellar masses) are required to measure the circularly averaged mass profile of a lens and its shape. \Glass\ uses an under-constrained adaptive grid of mass pixels to model the lens, searching through thousands of models to marginalise over model uncertainties. Our key findings are as follows: (i) for pure lens data, multiple sources with wide redshift separation give the strongest constraints as this breaks the well-known mass-sheet or steepness degeneracy; (ii) a single quad with time delays also performs well, giving a good recovery of both the mass profile and its shape; (iii) stellar masses -- for lenses where the stars dominate the central potential -- can also break the steepness degeneracy, giving a recovery almost as good as having time delay data or multiple source redshifts; (iv) stellar kinematics are similarly powerful if the data extend beyond the half light radius of the stars $r \gg r_{1/2}$ (required to integrate out the effect of the unknown velocity anisotropy, $\beta(r)$); and (v) if the lensing data already probe the mass profile over the region $r \sim r_{1/2}$, then stellar kinematic data can be used to probe $\beta(r)$ -- an interesting dynamic quantity in its own right. Where information on the mass distribution from lensing and other probes becomes redundant, this opens up the possibility of using strong lensing to constrain cosmological models. We will study this, and present the first results from \Glass\ applied to real data, in forthcoming papers.
{\bf JR. Need to highlight also the shape recovery results here.} 
\end{abstract}

@@

@@

\subsection{Stellar kinematics}

@@

\Glass\ uses an adaptive grid of mass pixels
to model the lens, generating thousands of models to marginalise over
uncertainties. It shares some heritage with an earlier code, \PixeLens\
\citep{1997MNRAS.292..148S,2008ApJ...679...17C}, but differs in several
important respects:

\begin{enumerate} 
\item it uses a significantly improved sampling strategy that is both faster and results in significantly smaller errors \citep{2012MNRAS.425.3077L};
\item it is built on a fully modular framework that can support different basis functions;
\item non-linear constraints can be applied in post-processing (for example, filters to remove models with spurious additional images or models inconsistent with stellar dynamics data are already implemented);
\item the stellar mass (if known) can be applied as a prior; and
\item adaptive resolution can be used in regions of interest. 
\end{enumerate}

This new functionality allows us to include non-lensing data constraints. Linear constraints -- like the stellar
mass -- are applied simultaneously with the strong lensing constraints;
non-linear constraints are applied in a post-processing step, where models are
rejected using a likelihood analysis. This latter was prohibitively expensive
with \PixeLens\ as only up to $\sim$1000 models could be generated in a
reasonable time. With \Glass, we can routinely generate hundreds of thousands
of models such that we can afford to discard some if they are inconsistent with
independent data.

\Glass\ is an evolution of the concepts employed by the free form modelling tool
\PixeLens\ but completely rewritten in Python and C. The key improvements are:
%
\begin{enumerate}
  \setcounter{enumi}{0}
  \item A modular framework allows new priors to be added and modified easily.
\end{enumerate}
%
Each prior is a simple function that adds linear constraints that operate on either
a single lens object or the entire ensemble of objects. \Glass\ comes with a number
of useful priors, but a user can write their own in the input file.
%
\begin{enumerate}
  \setcounter{enumi}{1}
  \item The basis functions approximating a model can be changed. 
\end{enumerate}
%
\Glass\ currently describes the lens mass as a collection of pixels, but the code
has been designed to support alternative methods. In particular, it is planned
to develop a module using Bessel functions. This will require a new set of 
priors that operate on these functions.
%
\begin{enumerate}
  \setcounter{enumi}{2}
  \item Non-linear constraints can be imposed in an automated post-processing step. 
\end{enumerate}
%
Once \Glass\ has generated an ensemble of models given the linear constraints, any number
of post processing functions can be applied. Not only can these functions be used to
derive new quantities from the mass models, they can also be used as a filter to 
accept or reject a model based on some non-linear constraint. The plotting functions
within \Glass\ will correctly display models that have been accepted or rejected.
%
\begin{enumerate}
  \setcounter{enumi}{3}
  \item The central region can be have a higher resolution to capture steep models. 
\end{enumerate}
%
By default, the mass distribution of the lens is described by uniform grid. However,
in the central region of a lensing galaxy where the mass profile may rise steeply,
a higher resolution can be specified that only applies to the inner pixels. These
pixels are subdivided into a smaller grid.
%
\begin{enumerate}
  \setcounter{enumi}{4}
  \item Stellar density can be used as an additional constraint.  
\end{enumerate}
%
The mass in inner regions of galaxies is often dominated by the stellar component
which one can estimate using standard mass-to-light models. This data can be added
to the potential as described later in \secref{stellar mass}. By using the stellar
mass one can place a lower bound on the mass and help constrain the inner most
mass profile.
%
\begin{enumerate}
  \setcounter{enumi}{5}
  \item Point or extended mass objects can be placed in the field.
\end{enumerate}
%
\PixeLens\ only allows for a shear term to be added to the potential (as shown
later in \eqnref{shear}) to account for mass external to model region. This
is useful to capture the gross effects of a distant neighbour. \Glass\ also
has a shear term but additional allows for additional analytic potentials to
be included. This can be used to model substructure or multiple neighbors close
to the main lens. The substructure may have only a small effect if the lens is
a single galaxy, but if the lens is group then a potential can be added for
each of the galaxies.
%
\begin{enumerate}
  \setcounter{enumi}{5}
  \item A new uniform sampling algorithm for high dimensional spaces.
\end{enumerate}
%
At the heart of \Glass\ lies a new algorithm for sampling the high dimensional
linear space that represents the modelling solution space. This algorithm was
described and tested in \cite{}. \Glass\ is also multi-threaded allowing it to
run efficiently on many-cored machines.  The software is freely available for
download. One interesting aspect is that the input files are themselves Python
programs. This allows a large amount of sophistication in setting up a lens at
runtime and also allows \Glass\ to be used as an external library from another
user written program.

\subsection{Discrete models}

For this paper we will restrict ourselves to using the pixelated basis set that
has been employed in previous work and is implemented in \PixeLens. Here we
will briefly review the main ideas. The algorithm for generating models in
\Glass\ samples a convex polytope in a high dimensional space whose interior
points are solutions to the lens equation and satisfy other physically
motivated linear priors (the sampling algorithm in \Glass\ is not the same as
employed in \PixeLens).  We therefore formulate all of our equations as
equations linear in the unknowns. We describe the density distribution $\kappa$
as a set of discrete grid cells or pixels $\kappa_i$ and rewrite the potential
(\eqnref{lensing potential}) as
%
\begin{equation}
  \psi(\vec\theta) = \sum_n \kappa_n Q_n(\vec\theta)
  \label{discrete potential}
\end{equation}
%
where the sum runs over all the pixels and $Q_n$ is the integral of the logarithm
over pixel $n$. The exact form for $Q$ is described in \appref{Q derivation}.
We can find the discretized lens equation by simply taking the gradient of the
above equations.

@@

\subsection{Removing models with extra images}

\subsection{A post-processing module for stellar kinematics}

@@

\subsection{Raytracing}
\label{Raytracing}
\Glass\ can also determine the position of images and time delays from 
particle-based simulation output given a source position $\vec\beta$. This is
used to generate the lens configurations used in the parameter study.  The
particles are first projected onto a very high resolution grid representing the
lens plane. The centers $\vec\theta_i$ of each of the grid cells are mapped
back onto the source plane using \eqnref{eqn:lens_equation}. If the location on
the source plane $\vec\beta_i$ is within a user specified
$\eps_\mathrm{accept}$ of $\vec\beta$ then $\vec\theta_i$ is 
accepted and further refined using a root finding algorithm until the distance
to $\vec\beta$ is nearly zero. If multiple points converge to an
$\eps_\mathrm{root}$ of each other then only one point is taken.  Care must be
taken that the grid resolution is high enough that the resulting image position
error is below the equivalent observational error. Time delays are then
calculated in order from the arrival time at each image (\eqnref{tau}).

@@

We now present a study of four mock galaxies with known analytic forms to
confirm that \Glass is able to correclty recover the mass profile.

>>>>

\begin{abstract}
We use a new non-parametric gravitational -- \Glass -- to determine what quality of data (strong lensing, stellar kinematics, and/or stellar masses) are required to measure the circularly averaged mass profile of a lens and its shape. \Glass\ uses an under-constrained adaptive grid of mass pixels to model the lens, searching through thousands of models to marginalise over model uncertainties. Our key findings are as follows: (i) for pure lens data, multiple sources with wide redshift separation give the strongest constraints as this breaks the well-known mass-sheet or steepness degeneracy; (ii) a single quad with time delays also performs well, giving a good recovery of both the mass profile and its shape; (iii) stellar masses -- for lenses where the stars dominate the central potential -- can also break the steepness degeneracy, giving a recovery almost as good as having time delay data or multiple source redshifts; (iv) stellar kinematics are similarly powerful if the data extend beyond the half light radius of the stars $r \gg r_{1/2}$ (required to integrate out the effect of the unknown velocity anisotropy, $\beta(r)$); and (v) if the lensing data already probe the mass profile over the region $r \sim r_{1/2}$, then stellar kinematic data can be used to probe $\beta(r)$ -- an interesting dynamic quantity in its own right. Where information on the mass distribution from lensing and other probes becomes redundant, this opens up the possibility of using strong lensing to constrain cosmological models. We will study this, and present the first results from \Glass\ applied to real data, in forthcoming papers.
\end{abstract}

@@

{\bf Is ``Schneider's degeneracy" something new? Should we cite it here? -- {\tt http://adsabs.harvard.edu/abs/2013arXiv1306.4675S}}

@@

For many galaxy lenses, the central potential is dominated by the visible matter. In this situation, constraints on the stellar mass distribution are invaluable. \citep{2008MNRAS.383..857F} demonstrate that with just photometric data in XXX bands, and an assumed initial mass function for the stars, they are able to reliably constrain the stellar mass to XXX\% accuracy. This gives a constraint on the full lensing mass map:

\begin{equation} 
M(\vec\theta) > M_*(\vec\theta)
\end{equation} 
where $M_*$ is the stellar mass (see also similar work by ZZZZZ). Given the above work, we assume from here on that the stellar mass for our mock N-body data can be reliably determined and used as a lower-bound constraint on our lens models.

{\bf JR. Prasenjit: you know more than me about this. Can you fill in the XXX's? We should probably also say a bit more about Ignacio's methodology and why the errors are so good etc etc.}

\subsection{Stellar kinematics}\label{sec:kinematics}

Another useful constraint follows from the velocity of stars within the lensing galaxy. Assuming spherical symmetry, stars obey the projected Jeans equations \citep[e.g.][]{2008gady.book.....B}:

\begin{equation}
\sigma_p^2(R) = \frac{2}{I(R)}\int_R^\infty dr\nu(r)f(r)\frac{GM(r)}{r}
\label{eqn:sphericaljeans}
\end{equation}
where $\sigma_p$ is the projected velocity dispersion of the stars as a function of projected radius $R$; $I(R)$ is the surface density of the stars; $\nu(r)$ is the three dimensional stellar density; $f(r)$ is a function that depends only on the velocity anisotropy $\beta(r) = 1 - \sigma_t^2/\sigma_r^2$, where $\sigma_{t,r}$ are the tangential and radial velocity dispersions respectively:

\begin{equation}
f(r) = XXX;
\end{equation} 
$G$ is Newton's gravitational constant; and $M(r)$ is the mass profile that we would like to measure.

It is immediately clear from \eqnref{eqn:sphericaljeans} that, even assuming spherical symmetry, we have a degeneracy between the enclosed mass profile $M(r)$ and the velocity anisotropy $\beta(r)$. This can be understood intuitively since $\beta(r)$ measures the relative importance of radial versus circular orbits and is intrinsically difficult to constrain given only one component of the velocity vector for each star. Nonetheless, $\beta(r)$ can be constrained given sufficiently many stars, since radial Doppler velocities sample eccentric orbits as $r\rightarrow 0$ and tangential orbits as $r\rightarrow \infty$ (CITES). It can also be estimated if an independent measure of $M(r)$ is available -- for example coming from strong lensing. Finally, we may integrate out the effect of unknown $\beta(r)$ by performing a luminosity weighted average over all radii:

\begin{equation}
\overline{\sigma} = \frac{\int_0^\infty I(R) \sigma_p^2(R) 2\pi R dR}{\int_0^\infty I(R) 2\pi R dR}
\label{eqn:virialaverage}
\end{equation} 
This produces a single robust measure of the mass $\overline{\sigma}$ that is independent of $\beta(r)$, provided that the stellar system is in a dynamic quasi-equilibrium that obeys the virial theorem (CITES). In practice, we need not perform the above integral to infinity (this is impractical for real data since the surface brightness at large $R$ will eventually fall below the detection limit). Rather, we should see \eqnref{eqn:virialaverage} as a means to understand why fitting the projected velocity dispersion profile of stars gives us a single robust mass measure that is independent of $\beta(r)$ at some critical radius, typically $\sim$ the half light radius of the stellar distribution ($r_{1/2}$; CITES).

We describe our numerical solution of \eqnref{eqn:sphericaljeans} in \S\ref{sec:glasskinematics} and present tests applied to mock data in \S\ref{sec:results}.

@@

\Glass\ is an evolution of the concepts employed by the free form modelling tool
\PixeLens\ but completely rewritten in Python and C. The key improvements are:
%
\begin{enumerate}
  \setcounter{enumi}{0}
  \item A modular framework allows new priors to be added and modified easily.
\end{enumerate}
%
Each prior is a simple function that adds linear constraints that operate on either
a single lens object or the entire ensemble of objects. \Glass\ comes with a number
of useful priors, but a user can write their own in the input file.
%
\begin{enumerate}
  \setcounter{enumi}{1}
  \item The basis functions approximating a model can be changed. 
\end{enumerate}
%
\Glass\ currently describes the lens mass as a collection of pixels, but the code
has been designed to support alternative methods. In particular, it is planned
to develop a module using Bessel functions. This will require a new set of 
priors that operate on these functions.
%
\begin{enumerate}
  \setcounter{enumi}{2}
  \item Non-linear constraints can be imposed in an automated post-processing step. 
\end{enumerate}
%
Once \Glass\ has generated an ensemble of models given the linear constraints, any number
of post processing functions can be applied. Not only can these functions be used to
derive new quantities from the mass models, they can also be used as a filter to 
accept or reject a model based on some non-linear constraint. The plotting functions
within \Glass\ will correctly display models that have been accepted or rejected.
%
\begin{enumerate}
  \setcounter{enumi}{3}
  \item The central region can be have a higher resolution to capture steep models. 
\end{enumerate}
%
By default, the mass distribution of the lens is described by uniform grid. However,
in the central region of a lensing galaxy where the mass profile may rise steeply,
a higher resolution can be specified that only applies to the inner pixels. These
pixels are subdivided into a smaller grid. {\bf JR. I think Jonathan is now applying this by default. Is that right Jonathan?} 
%
\begin{enumerate}
  \setcounter{enumi}{4}
  \item Stellar density can be used as an additional constraint.  
\end{enumerate}
%
The mass in inner regions of galaxies is often dominated by the stellar component
which one can estimate using standard mass-to-light models. This data can be added
to the potential as described later in \secref{stellar mass}. By using the stellar
mass one can place a lower bound on the mass and help constrain the inner most
mass profile.
%
\begin{enumerate}
  \setcounter{enumi}{5}
  \item Point or extended mass objects can be placed in the field.
\end{enumerate}
%
\PixeLens\ only allows for a shear term to be added to the potential (as shown
later in \eqnref{shear}) to account for mass external to model region. This
is useful to capture the gross effects of a distant neighbour. \Glass\ also
has a shear term but additional allows for additional analytic potentials to
be included. This can be used to model substructure or multiple neighbors close
to the main lens. The substructure may have only a small effect if the lens is
a single galaxy, but if the lens is group then a potential can be added for
each of the galaxies.
%
\begin{enumerate}
  \setcounter{enumi}{5}
  \item A new uniform sampling algorithm for high dimensional spaces.
\end{enumerate}
%
At the heart of \Glass\ lies a new algorithm for sampling the high dimensional
linear space that represents the modelling solution space. This algorithm was
described and tested in \cite{2012MNRAS.425.3077L}. \Glass\ is also multi-threaded allowing it to
run efficiently on many-cored machines.  The software is freely available for
download. One interesting aspect is that the input files are themselves Python
programs. This allows a large amount of sophistication in setting up a lens at
runtime and also allows \Glass\ to be used as an external library from another
user written program.

%\Glass\ uses an adaptive grid of mass pixels
%to model the lens, generating thousands of models to marginalise over
%uncertainties. It shares some heritage with an earlier code, \PixeLens\
%\citep{1997MNRAS.292..148S,2008ApJ...679...17C}, but differs in several
%important respects: 
%\begin{enumerate} 
%\item it uses a significantly improved sampling strategy that is both faster and results in significantly smaller errors \citep{2012MNRAS.425.3077L};
%\item it is built on a fully modular framework that can support different basis functions;
%\item non-linear constraints can be applied in post-processing (for example, filters to remove models with spurious additional images or models inconsistent with stellar dynamics data are already implemented);
%\item the stellar mass (if known) can be applied as a prior; and
%\item adaptive resolution can be used in regions of interest. 
%\end{enumerate}
%This new functionality allows us to include non-lensing data constraints. Linear constraints -- like the stellar
%mass -- are applied simultaneously with the strong lensing constraints;
%non-linear constraints are applied in a post-processing step, where models are
%rejected using a likelihood analysis. This latter was prohibitively expensive
%with \PixeLens\ as only up to $\sim$1000 models could be generated in a
%reasonable time. With \Glass, we can routinely generate hundreds of thousands
%of models such that we can afford to discard some if they are inconsistent with
%independent data.

\subsection{Discrete models}\label{sec:discrete}
For this paper we will restrict ourselves to using the pixelated basis set that
has been employed in previous work and is implemented in \PixeLens. Here we
will briefly review the main ideas. The algorithm for generating models in
\Glass\ samples a convex polytope in a high dimensional space whose interior
points are solutions to the lens equation and satisfy other physically
motivated linear priors (the sampling algorithm in \Glass\ is not the same as
employed in \PixeLens).  We therefore formulate all of our equations as
equations linear in the unknowns. We describe the density distribution $\kappa$
as a set of discrete grid cells or pixels $\kappa_i$ and rewrite the potential
(\eqnref{lensing potential}) as
%
\begin{equation}
  \psi(\vec\theta) = \sum_n \kappa_n Q_n(\vec\theta)
  \label{discrete potential}
\end{equation}
%
where the sum runs over all the pixels and $Q_n$ is the integral of the logarithm
over pixel $n$. The exact form for $Q$ is described in \appref{Q derivation}.
We can find the discretized lens equation by simply taking the gradient of the
above equations.

@@

\subsection{Raytracing}\label{Raytracing}
\Glass\ can also determine the position of images and time delays from 
particle-based simulation output given a source position $\vec\beta$. This is
used to generate the lens configurations used in the parameter study.  The
particles are first projected onto a very high resolution grid representing the
lens plane. The centers $\vec\theta_i$ of each of the grid cells are mapped
back onto the source plane using \eqnref{eqn:lens_equation}. If the location on
the source plane $\vec\beta_i$ is within a user specified
$\eps_\mathrm{accept}$ of $\vec\beta$ then $\vec\theta_i$ is 
accepted and further refined using a root finding algorithm until the distance
to $\vec\beta$ is nearly zero. If multiple points converge to an
$\eps_\mathrm{root}$ of each other then only one point is taken.  Care must be
taken that the grid resolution is high enough that the resulting image position
error is below the equivalent observational error. Time delays are then
calculated in order from the arrival time at each image (\eqnref{tau}).

\subsection{Removing models with extra images}\label{sec:glassextraimages} 
While linear constraints can be applied on-the-fly in \Glass, non-linear constraints must be applied in post-processing. Models that are inconsistent with such constraints must then be statistically discarded via a likelihood analysis. An example of such a non-linear constraint is the spurious presence of unobserved images. This `null-space' prior was first proposed and explored by \citet{2006MNRAS.367.1209L} and found to be extremely powerful. We find that our gradient prior in \Glass\ (see \S\ref{sec:discrete}), performs much of the same function as \citeauthor{2006MNRAS.367.1209L}'s null-space prior, but some models can still rarely turn up spurious images. We reject these in a post-processing step, where we sweep through the model ensemble applying the ray tracing algorithm described in \ref{Raytracing}. Models are accepted or rejected according to the criteria:

\begin{equation}
XXX
\end{equation} 
Typically, this results in a rejection rate of XXX\%.

\subsection{A post-processing module for stellar kinematics}\label{sec:glasskinematics} 
Similarly to the null-space constraint (\S\ref{sec:glassextraimages}), stellar kinematic constraints constitute a non-linear prior on the mass map and must be applied in post-processing. We sweep through the model ensemble performing an Abel deprojection to determine $M(r)$ from the projected surface density assuming spherical symmetry \citep[e.g.][]{2008gady.book.....B}:

\begin{equation} 
M(r) = XXX
\end{equation} 
This de-projection algorithm was tested on triaxial figures in \citet{2006ApJ...652L...5S}. They found that for triaxialities typical of our current cosmology, the method works extremely well unless the triaxial figure is projected directly along the line of sight such that we see the galaxy or galaxy cluster `down the barrel'. Such a situation is unlikely, but in any case avoidable since the resultant figure appears spherical in projection. This leads to the seemingly counter-intuitive result that the kinematic constraints -- that rely on the above de-projection -- are most secure for systems that do not appear spherical in projection (unless independent data can confirm the three dimensional shape is indeed very round).

We use the deprojected mass to numerically solve equation \ref{eqn:sphericaljeans} for constant $\beta(r)$, assuming either $\beta(r) = 1$ or $\beta(r) = 0$ at all radii to bracket the two extremum situations. Where the data are good enough, these two may be distinguished giving dynamical information about $\beta(r)$. In more typical situations, however, we seek to simply marginalise over the effect of $\beta(r)$, using the stellar kinematics as a robust measure of $M(r_{1/2})$ (see \S\ref{sec:kinematics}).

@@

@@

We now present a study of four mock galaxies with known analytic forms. These are used to
verify that \Glass\ is able to correctly recover the mass profile, and -- more importantly -- to determine what type and quality of data best constrain the mass profile and shape of a lens.
